
# Packages installation
```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

#BiocManager::install("AnnotationDbi")
#BiocManager::install("org.Hs.eg.db")
BiocManager::install("org.Mm.eg.db")
install.packages('viridis')
BiocManager::install('mixOmics')
install.packages('RColorBrewer')
#BiocManager::install("GO.db")
#BiocManager::install("systemPipeR")
#BiocManager::install("clusterProfiler")
BiocManager::install("Glimma")
BiocManager::install("edgeR")
BiocManager::install("biomaRt")
```

# Packages loading
```{r}
#library("AnnotationDbi")
#library("biomaRt")
#library("org.Hs.eg.db")
library('org.Mm.eg.db')
library(edgeR)
library(limma)
library(Glimma)
library(gplots)
library(RColorBrewer)
```

# Differential gene expression (edgeR)
```{r}
# Wczytanie danych do R
seqdata <- read.table("DE/GSE60450_Lactation-GenewiseCounts.txt", stringsAsFactors = FALSE, header = T)

dim(seqdata)

sampleinfo <- read.delim("DE/SampleInfo.txt", header = T, stringsAsFactors = T) # Add `stringsAsFactors = TRUE`, otherwise the columns will be treated as characters and will not be recognized as vectors

sampleinfo
```

## DGEList object creation
```{r}
y <- DGEList(counts = seqdata[, 3:14], genes = seqdata[,1], group = sampleinfo$CellType)
```

```{r}
names(y)
y$samples
```

## Annotation
```{r}
library(org.Mm.eg.db)
egSYMBOL <- toTable(org.Mm.egSYMBOL)
head(egSYMBOL)
m <- match(y$genes$genes, egSYMBOL$gene_id)
y$genes$Symbol <- egSYMBOL$symbol[m]
head(y$genes)
colnames(y) <- substr(colnames(y),start=1,stop=7)
head(y)
```

## Sample names
```{r}
table(colnames(y)==sampleinfo$SampleName)
```

## Data filtration
```{r}
d <- is.na(y$genes$Symbol)
y <- y[!d,]
nrow(y)
nrow(d)
```

```{r}
keep <- filterByExpr(y, min.count = 5, min.prop = 0.5)
table(keep)
```

```{r}
y <- y[keep,  keep.lib.sizes=FALSE]
```

### recalculating the size of libraries
```{r}
y$samples$lib.size <- colSums(y$counts)
y$samples
```

## TMM normalization 
```{r}
y1 <- calcNormFactors(y)
y1$samples
```

## Library sizes - plots
```{r}
y1$samples$lib.size
barplot(y1$samples$lib.size,names=colnames(y1),las=2)
# Add a title to the plot
title("Barplot of library sizes")
```

## Distribution plots
```{r fig.height=4, fig.width=8}
par(mfrow=c(1,2))

logcounts <- cpm(y,log=TRUE)
boxplot(logcounts, xlab="", ylab="Log2 counts per million",las=2)
abline(h=median(logcounts),col="blue") # Adding a median line with logCPM
title("Boxplots of logCPMs (unnormalised)")

logcounts <- cpm(y1,log=TRUE)
boxplot(logcounts, xlab="", ylab="Log2 counts per million",las=2)
abline(h=median(logcounts),col="blue") # Adding a median line with logCPM
title("Boxplots of logCPMs (normalised)")
```

## MD plots (MAplot)
```{r fig.height=4, fig.width=8}
par(mfrow=c(1,2))

plotMD(cpm(y, log=TRUE), column=7)
abline(h=0, col="red", lty=2, lwd=2)

plotMD(cpm(y, log=TRUE), column=11)
abline(h=0, col="red", lty=2, lwd=2)
```

```{r fig.height=4, fig.width=8}
par(mfrow=c(1,2))

plotMD(y1,column = 7)
abline(h=0,col="grey")

plotMD(y1,column = 11)
abline(h=0,col="grey")
```

## Multidimensional scaling plots (MDSplots)
```{r}
par(mfrow=c(1,2))
# choosing a color scheme for the variable CellType
levels(sampleinfo$CellType)

col.cell <- c("purple","orange")[sampleinfo$CellType]
data.frame(sampleinfo$CellType,col.cell)
# creating an MDS plot for CellType
plotMDS(y1,col=col.cell)
# Adding the legend
legend("topleft",fill=c("purple","orange"),legend=levels(sampleinfo$CellType))
# Adding the title
title("Cell type")

# the same for "Status" variable
levels(sampleinfo$Status)
col.status <- c("blue","red","green")[sampleinfo$Status]
col.status
plotMDS(y1,col=col.status, pch = 5)
legend("topleft",fill=c("blue","red","green"),legend=levels(sampleinfo$Status),cex=0.8)
title("Status")
```

## Preparing the experimental design matrix
```{r}
#group <- paste(sampleinfo$CellType, sampleinfo$Status, sep=".")
#group <- factor(group)
#group
########## for other comparisons ########## 
group = sampleinfo$CellType
group <- factor(group)
group
```

```{r}
design <- model.matrix(~ 0 + group)
colnames(design) <- levels(group)
design
```

## Estimation of dispersion
```{r}
y1 <- estimateDisp(y1, design, robust=TRUE)
y1$common.dispersion
plotBCV(y1)
```

## Fitting a linear model
```{r}
fit <- glmQLFit(y1, design, robust=TRUE)
head(fit$coefficients)
```

## Selection of comparisons for testing
```{r}
cont.matrix <- makeContrasts(B.PregVsLac = basal.pregnant - basal.lactate, levels = design)
cont.matrix <- makeContrasts(BVsL = basal - luminal, levels = design)

#cont.matrix <- makeContrasts(B.PregVsLac = basal.pregnant - basal.lactate, L.PregVsLac = luminal.pregnant - luminal.lactate, levels = design)

cont.matrix
```

## Testing procedure
```{r}
qlf <- glmQLFTest(fit, contrast = cont.matrix)
topTags(qlf)
summary(decideTests(qlf))
```

## Testing procedure including FC (fold change)
```{r}
tr <- glmTreat(fit, contrast=cont.matrix, lfc=log2(1.5))
topTags(tr)
summary(decideTests(tr))
```

## Plots After Differential Analysis

### MAplot
```{r fig.height=4, fig.width=8}
par(mfrow=c(1,2))

plotMD(qlf)
abline(h=c(-1,1), col="blue")

plotMD(tr)
abline(h=c(-1,1), col="blue")
```

### Data for the volcano plot
```{r}
FDR <- p.adjust(qlf$table$PValue, method="BH")

volcanoData <- data.frame(
  logFC = qlf$table$logFC.B.PregVsLac,
  negLogPval = -log10(FDR)
  )

volcanoData2 <- data.frame(
  logFC = qlf$table$logFC.L.PregVsLac,
  negLogPval = -log10(FDR)
  )

head(volcanoData)
head(volcanoData2)
```

### volcano plot
```{r}
library(ggplot2)

ggplot(volcanoData, aes(logFC, negLogPval)) +
geom_point()

ggplot(volcanoData2, aes(logFC, negLogPval)) +
geom_point()
```

```{r}
library(dplyr)
#volcanoData <- volcanoData %>%
#  mutate(top50 = ifelse(row_number(desc(negLogPval)) < 51, T, F))

volcanoData <- volcanoData %>%
  mutate(sig = ifelse(volcanoData$negLogPval > 1.30103, T, F))

ggplot(volcanoData, aes(logFC, negLogPval, color = sig)) +
  geom_point() + theme(legend.position = "none") +
  scale_color_manual(values = c("black", "blue"))
```

### Heatmap  (FDR < 0.05)
```{r}
library(viridis)
library(mixOmics)
library(RColorBrewer)
```

```{r}
cpm_y <- cpm(tr, log=TRUE, prior.count = 1)
head(cpm_y)
```

```{r include=FALSE}
table = topTags(tr, n=Inf)
table = as.data.frame(table)
selY <- cpm_y[rownames(table)[table$FDR<0.01 & abs(table$logFC)>2.5],]
```

```{r fig.height=10, fig.width=10}
cimColor <- turbo(255, begin = 0.1, end = 0.9, direction = 1)
cim(t(selY), color = cimColor, symkey=F, transpose = T)
```

# Saving the results
```{r}
# generating results for all genes
edgeR.res <- topTags(tr, sort.by = "p", n = "Inf")

# saving the results to a file
write.csv(edgeR.res,
	file = "B.PregVsLacResults.csv",
	row.names=FALSE)
```

# End
```{r}
# packages info
sessionInfo()
```